---
title: "End of Semester Exam"
author: "Nathan Orang'o"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# Libraries Used in this Document
library(ggplot2)
library(dplyr)
```

# A recent national study showed that approximately 44.7% of college students have used Wikipedia as a source in at least one of their term papers. Let X equal the number of students in a random sample of size n = 31 who have used Wikipedia as a source.

## 1. [How is X distributed?]{.underline}

-   X follows a `Binomial Distribution`, for the following reasons
    1.  There is a fixed number of trials (31).
    2.  Each trial has only two possible outcomes (Used Wikipedia or did not use Wikipedia).
    3.  The probability of success (using Wikipedia) is the same for each trial.

## 2. [Generate and plot the probability mass function.]{.underline}

### [What is a Probability Mass Function]{.underline}

-   The `Probability Mass Function (PMF)` quantifies the probability that a discrete random variable(X -\> The number of students out of the sample who used Wikipedia) equals a specific value.
-   This binomial distribution's PMF is given by the formula:

$$
P(X=k) = \binom{n}{k} p^k (1-p)^{n-k}
$$

Where:

1.  $P(X=k)$ is the probability of exactly $k$ students having used Wikipedia as a source.
2.  $n = 31$ is the total number of trials (students in the sample).
3.  $k$ represents the number of successes (students using Wikipedia), ranging from 0 to $n$.
4.  $p = 0.447$ is the probability of a single trial resulting in success (a student using Wikipedia).
5.  $\binom{n}{k}$ is the binomial coefficient, calculated as $\frac{n!}{k!(n-k)!}$, representing the number of ways to choose $k$ successes out of $n$ trials.

-   This calculated PMF enables us to calculate and visualize the likelihood of observing any specific number of students (from 0 to 31) using Wikipedia in their term papers, for example, vizualising the likelihood of finding 17 students who used Wikipedia as a source in their term papers.
-   Below is a plot of the binomial distribution using the `dbinom` function;

```{r}
n <- 31
p <- 0.447
x_values <- 0:n
pmf_values <- dbinom(x_values, size=n, prob=p)

plot(x_values, pmf_values, type="h", lwd=7, col="blue", 
     main="PMF of Binomial Distribution", xlab="Number of Students", ylab="Probability")
```

### [Explanation of the code]{.underline}

-   The code snippet `n <- 31` assigns the total number of students in the sample(31) to `n`.
-   The code snippet `p <- 0.447` assigns the probability that any given student has used Wikipedia to `p`.
-   The code snippet `x_values <- 0:n` assigns a vector containing all possible numbers of students using Wikipedia to `x_values`.
-   The code snippet `pmf_values <- dbinom(x_values, size=n, prob=p)` calculates the PMF values for each count of successes using the `dbinom` function.
-   The code snippet `# Plot the PMF plot(x_values, pmf_values, type="h", lwd=2, col="blue",main="PMF of Binomial Distribution", xlab="Number of Students", ylab="Probability")` uses the plot function is used to create a visual representation of the PMF. The arguments specify that it should be a bar plot `(type="h")` with blue lines `(col="blue")` of width 7 `(lwd=7)`. The plot is labeled with a main title and axis labels indicating what the axes represent.

### [Description of the Plot]{.underline}

-   The x-axis represents the number of students, ranging from 0 to 31, who have used Wikipedia as a source.
-   The y-axis indicates the probability associated with each number of students on the x-axis. The scale of this axis goes from 0 to a maximum that's slightly above the highest probability observed in the data.
-   Each vertical blue line represents the probability of that exact number of students using Wikipedia as per the binomial distribution with the specified parameters. The height of each bar corresponds to the PMF value calculated for each $k$ value (number of students).
-   The distribution of the lines appears to be symmetric around its mean, with the highest bars (probabilities) concentrated around the center of the plot (This is typical for a binomial distribution where $p$ is not too close to 0 or 1).

## 3. [Generate and plot the cumulative distribution function]{.underline}

### [What is Cumulative Distribution Function]{.underline}

-   The `Cumulative Distribution Function (CDF)` for a discrete random variable gives the probability that the variable(X) is less than or equal to a certain value.
-   Here, it is used in conjunction with the binomial distribution to determine probability that the number of students who have used Wikipedia as a source is within a certain range.
-   The CDF is calculated as The Cumulative Distribution Function (CDF) for a discrete random variable is defined as:

$$ F(k;n,p) = P(X \leq k) = \sum_{i=0}^{k} \binom{n}{i} p^i (1-p)^{n-i} $$

Where:

-   $F(k;n,p)$ is the value of the CDF at $k$.

-   $P(X \leq k)$ is the probability that the random variable $X$ is less than or equal to $k$.

-   $\binom{n}{i}$ is the binomial coefficient for $i$ successes in $n$ trials.

-   $p^i$ is the probability of having exactly $i$ successes.

-   $(1-p)^{n-i}$ is the probability of the remaining $n-i$ trials being failures.

```{r}
# Defining the parameters for the binomial distribution
n <- 31        # number of trials
p <- 0.447     # probability of success

# Generating a sequence of possible success counts from 0 to 31
x_values <- 0:n

# Calculating the cumulative distribution function (CDF) values
cdf_values <- pbinom(x_values, size=n, prob=p)

# Plotting the CDF
plot(x_values, cdf_values, type="s", col="red", lwd=2, 
     main="CDF of Binomial Distribution", xlab="Number of Students", ylab="Cumulative Probability",
     ylim=c(0, 1))
```

### [Explanation of the Code]{.underline}

-   The code snippet `n <- 31 p <- 0.447` sets up the n and p parameters for the binomial distribution, where n is the number of independent trials (or students in the context of the question), and p is the probability of a single trial resulting in success (a student using Wikipedia as a source).
-   The code snippet `x_values <- 0:n` assigns a sequence of integers from 0 to n (31), represent all possible numbers of students that could have used Wikipedia, from none (0) to all (31) to `x_values`.
-   The code snippet `cdf_values <- pbinom(x_values, size=n, prob=p)` calculates the CDF for the binomial distribution at each value in x_values using the `pbinom` function.
-   The code chunk `plot(x_values, cdf_values, type="s", col="red", lwd=2,  main="CDF of Binomial Distribution", xlab="Number of Students", ylab="Cumulative Probability", ylim=c(0, 1))` uses the `plot` function to generate a graph of the CDF.

### [Description of the Plot]{.underline}

-   The x-axis represents the number of students, ranging from 0 to 31, corresponding to the possible number of students who might have used Wikipedia as a source.
-   The y-axis represents cumulative probability, which ranges from 0 to 1,describing the probability of having up to a certain number of students who have used Wikipedia.
-   For any point on this curve, the y-value gives the probability that a random variable X (the number of students who have used Wikipedia) is less than or equal to that number of students. As you moving along the x-axis, the steps rise cumulatively, reflecting the increasing probability of encountering at least that many students who have used Wikipedia.

## 4. [Find the probability that X is]{.underline}

## (a) [Equal to 17]{.underline}

Using the binomial probability formula:

$$ P(X = k) = \binom{n}{k} p^k (1-p)^{n-k} $$

Substitute $n = 31$, $k = 17$, and $p = 0.447$:

$$ P(X = 17) = \binom{31}{17} \times 0.447^{17} \times (1 - 0.447)^{31 - 17} $$

As manual calculation involves a lengthy, inefficient process, I shall calculate the probabilities computationally using the `pbinom` function;

```{r}
# Defining parameters for the binomial distribution
n <- 31      
p <- 0.447     


# P(X = 17)

prob_a <- dbinom(17, n, p)
prob_a
```

-   The computed probability is approximately 0.07532248, indicating that there is a 7.53% chance that exactly 17 out of 31 students have used Wikipedia as a source in their term papers.

## (b) [At most 13]{.underline}

The cumulative probability up to $k$ is the sum of individual probabilities:

$$ P(X \leq k) = \sum_{i=0}^{k} \binom{n}{i} p^i (1-p)^{n-i} $$

Substitute $n = 31$, $k = 13$, and $p = 0.447$:

$$ P(X \leq 13) = \sum_{i=0}^{13} \binom{31}{i} \times 0.447^{i} \times (1 - 0.447)^{31 - i} $$

Calculating computationally;

```{r}
prob_b <- pbinom(13, n, p)
prob_b
```

-   The probability that no more than 13 students have used Wikipedia as a source in their term papers is 0.451337. This value represents the sum of probabilities for all instances where the number of students using Wikipedia ranges from 0 to 13.

## (c) [Greater than 11]{.underline}

This is the complement to the cumulative probability up to $k = 11$:

$$ P(X > 11) = 1 - P(X \leq 11) $$

Substituting the values:

$$ P(X > 11) = 1 - \sum_{i=0}^{11} \binom{31}{i} \times 0.447^{i} \times (1 - 0.447)^{31 - i} $$

Calculating the computationally;

```{r}
prob_c <- 1 - pbinom(11, n, p)
prob_c
```

-   The probability that more than 11 students have used Wikipedia is 0.8020339.

-   This result indicates that there is an approximately 80.20% chance that more than 11 out of 31 students have used Wikipedia as a source for their term papers.

## (d) [At least 15]{.underline}

As above, I used the complement to the cumulative probability up to $k = 14$:

$$ P(X \geq 15) = 1 - P(X \leq 14) $$

Substituting the values:

$$ P(X \geq 15) = 1 - \sum_{i=0}^{14} \binom{31}{i} \times 0.447^{i} \times (1 - 0.447)^{31 - i} $$

Calculating computationally;

```{r}
prob_d <- 1 - pbinom(14, n, p)
prob_d
```

-   The computed probability (0.406024), indicates there is approximately a 40.60% chance that at least 15 out of 31 students have used Wikipedia as a source for their term papers.

## (e) [Between 16 and 19, inclusive]{.underline}

For a range of values, we subtract the cumulative probability up to the lower limit from the cumulative probability up to the upper limit:

$$ P(16 \leq X \leq 19) = P(X \leq 19) - P(X \leq 15) $$

With our values:

$$ P(16 \leq X \leq 19) = \sum_{i=0}^{19} \binom{31}{i} \times 0.447^{i} \times (1 - 0.447)^{31 - i} - \sum_{i=0}^{15} \binom{31}{i} \times 0.447^{i} \times (1 - 0.447)^{31 - i} $$

Calculating Computationally;

```{r}
prob_e <- pbinom(19, n, p) - pbinom(15, n, p)
prob_e 
```

-   The computed probability 0.2544758) indicates there is approximately a 25.45% chance that between 16 and 19 students (inclusive) have used Wikipedia as a source for their term papers.

## 5. [Give the mean of X, denoted E(X).]{.underline}

The mean of the binomial distribution, denoted $E(X)$, is calculated as:

$$ E(X) = n \times p $$

Substituting for $n = 31$ and $p = 0.447$, this becomes:

$$ E(X) = 31 \times 0.447 $$

The mean of the binomial distribution therefore is $13.857$

This can also be calculated computationally;

```{r}
# Defining the parameters
n <- 31
p <- 0.447

# Calculating the mean of X
mean_X <- n * p

# Displaying the mean
mean_X
```

-   The computed mean (13.857) represents the expected number of students, on average, who have used Wikipedia as a source in their term papers, from a sample of 31 students.

-   The mean reflects the central tendency of the distribution, indicating that, under the binomial model, you can expect around 14 students in the sample to have used Wikipedia.

## 6. [Give the variance and standard deviation of X.]{.underline}

The variance of the binomial distribution is calculated as:

$$ \text{Var}(X) = n \times p \times (1 - p) $$

With $n = 31$ and $p = 0.447$, we have:

$$ \text{Var}(X) = 31 \times 0.447 \times (1 - 0.447) $$

Therefore, the variance is $7.662921$

The standard deviation is the square root of the variance:

$$ \text{SD}(X) = \sqrt{\text{Var}(X)} $$

Thus:

$$ \text{SD}(X) = \sqrt{7.662921)} $$

Therefore, the Standard Deviation is $2.768198$

The above can be calculated computationally;

```{r}
# Defining the parameters
n <- 31
p <- 0.447

# Calculating the variance of X
variance_X <- n * p * (1 - p)

# Calculating the standard deviation of X
sd_X <- sqrt(variance_X)

# Displaying the variance and standard deviation
variance_X
sd_X

```

-   The calculated variance (7.662921) signifies the degree to which the number of students using Wikipedia as a source deviates from the mean (13.857) in the sample of 31 students. It quantifies the spread (dispersion) of the binomial distribution around this mean value.

-   The calculated Standard Deviation (2.768198) provides a measure of this spread in the same units as the original data (number of students). This value indicates that, on average, the actual number of students using Wikipedia as a source in any given sample of 31 will typically vary by about 2.77 students from the mean of approximately 13.86 students.
